Towards a standardized evaluation of imputation methodology

### 2022-03-31 Formatted/Formatted.tex
Fix TODOs

Evaluating imputation methodology requires a missing data problem to be solved. After establishing a comparative truth from a data-generating mechanism, some form of missingness therefore has to be induced. Often, however, reports of simulation studies remain vague about the actual missingness conditions under investigation and, even worse, some authors only report something like:
\begin{quote}
\textit{We generated missing data following a missing at random missingness mechanism.}
\end{quote}
This should be considered unacceptable as claims about the validity of the imputation inference heavily depend on the simulated missingness conditions, such as missingness mechanisms and missingness patterns. Missingness mechanisms describe the relationship between missing entries and observed data values and are generally categorized into MCAR, MAR and MNAR mechanisms \citep[][see Table \ref{table:mech}]{rubi76}. [TODO: maybe remove mechanisms here?] Missingness patterns concern the location of missing entries across incomplete data \citep[][p. 8]{litt20}. Under this definition, there is a row-wise element to the missingness pattern describing which variables are jointly observed, and a column-wise element encompassing the amount of missingness in the data.

\begin{table}[tb]
\begin{center}
\caption{Missingness mechanisms. [TODO: add PMAR?]}
\label{table:mech}
\begin{tabular}{lll}
\hline
      & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & Missing completely at random: The probability to be missing is the same for all \\
      & cases. In other words, the missing values are missing at random and the observed \\
      & values are observed at random. \\
MAR   & Missing at random: The probability to be missing is the same within groups of \\
      & cases defined by the observed data only. The essence is that observed relations in \\
      & the data inform the missingness, such that MCAR may be assumed within the \\ 
      & observed groups. \\
MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
      & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
      & inferred from the observed data. This is considered non-ignorable missingness \\
      & \citep[see e.g.][]{rubi76}. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

The evaluation criteria used to assess imputation performance vary from one simulator to another. This is not surprising as people from different fields could have a different focus on the problem at hand. There are, however, some overarching issues with assessing imputation method performance. In the first place, diagnostic evaluation of the generated imputations is often left out of simulation study results. Problems with the imputation-generating process (e.g., an iterative imputation algorithm) may then be overlooked. In the second place, there are pitfalls in the evaluation of the estimates that are obtained by fitting the analysis model (i.e. the complete data model used to estimate the comparative truth). The choice of performance measures might [dictate (TODO: re-read and find better word!)] the results of the simulation study.

@@

First, one should always consider MCAR missingness (see Table \ref{table:mech} for definitions). This mechanism should be used as a reference condition. Every imputation method should yield valid imputations under MCAR. We also know that, under MCAR, the distribution of imputed values should be equivalent to the distribution of observed values for a given variable, which may offer an indication of imputation model (mis)fit. [TODO: leg uit!]

@@

Third, a non-ignorable or MNAR mechanism might fall outside many studies' scope, yet could yield informative insights for daily practice. Even if an imputation method is specifically developed with ignorable missingness in mind, chances are that the method will be applied to non-ignorable empirical missingness at some point in time. It would, therefore, be wise to include MNAR missingness in the simulation study just in case. Alternatively, sensitivity analyses may provide an indication of the validity of the obtained inference, given that the assumed missingness mechanism is suspected to be invalid \citep[see e.g.][part 5]{molenberghs2014handbook}.

Apart from the missingness mechanisms, the missingness pattern must be varied. [TODO: beter bruggetje?] Remember that missingness is only ignorable under MAR when the parameter of the data is distinct and a-priori independent from the parameter of the missing data process. Under MAR missingness we assume that we may use the observed data to make inferences about the joint (observed and unobserved) data. The dependency of the procedure on the assumption under which we obtain inference is only influenced by the amount of missingness. If there is no missingness--or if there is no data, for that matter--the inference does not depend on the assumption. Alternatively, the validity of assumptions becomes increasingly important when the missingness increases. Since we control the MAR mechanism, the assumption under which we may solve the missing data problem should hold and it is only fair to assess performance under stringent missingness conditions. We, therefore, propose to evaluate imputation methodology under 10\%, 25\% and 50\% univariate missingness (see Table \ref{table:prop}).

>>>>

Evaluating imputation methodology requires a missing data problem to be solved. After establishing a comparative truth from a data-generating mechanism, some form of missingness therefore has to be induced. Often, however, reports of simulation studies remain vague about the actual missingness conditions under investigation and, even worse, some authors only report something like:
\begin{quote}
\textit{We generated missing data following a missing at random missingness mechanism.}
\end{quote}
This should be considered unacceptable as claims about the validity of the imputation inference heavily depend on the simulated missingness conditions, such as missingness mechanisms and missingness patterns. Missingness mechanisms describe the relationship between missing entries and observed data values and are generally categorized into MCAR, MAR and MNAR mechanisms \citep[][see Table \ref{table:mech}]{rubi76}. Missingness patterns concern the location of missing entries across incomplete data \citep[][p. 8]{litt20}. Under this definition, there is a row-wise element to the missingness pattern describing which variables are jointly observed, and a column-wise element encompassing the amount of missingness in the data.

\begin{table}[tb]
\begin{center}
\caption{Missingness mechanisms.}
\label{table:mech}
\begin{tabular}{lll}
\hline
      & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & Missing completely at random: The probability to be missing is the same for all \\
      & cases. In other words, the missing values are missing at random and the observed \\
      & values are observed at random. \\
MAR   & Missing at random: The probability to be missing is the same within groups of \\
      & cases defined by the observed data only. The essence is that observed relations in \\
      & the data inform the missingness, such that MCAR may be assumed within the \\ 
      & observed groups. \\
MNAR  & Missing not at random: The probability to be missing depends on unobserved \\
      & aspects of the data. The cause of the missingness is unknown and cannot be \\ 
      & inferred from the observed data. This is considered non-ignorable missingness \\
      & \citep[see e.g.][]{rubi76}. \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

The evaluation criteria used to assess imputation performance vary from one simulator to another. This is not surprising as people from different fields could have a different focus on the problem at hand. There are, however, some overarching issues with assessing imputation method performance. In the first place, diagnostic evaluation of the generated imputations is often left out of simulation study results. Problems with the imputation-generating process (e.g., an iterative imputation algorithm) may then be overlooked. In the second place, there are pitfalls in the evaluation of the estimates that are obtained by fitting the analysis model (i.e. the complete data model used to estimate the comparative truth). The choice of performance measures might inadvertently influence simulation results.

@@

First, one should always consider MCAR missingness (see Table \ref{table:mech} for definitions). This mechanism should be used as a reference condition. Every imputation method must be able to yield valid imputations under MCAR, both in terms of distributional characteristics as well as statistical inference.

@@

Third, a non-ignorable or MNAR mechanism might fall outside many studies' scope, yet could yield informative insights for daily practice. Even if an imputation method is specifically developed with ignorable missingness in mind, chances are that the method will be applied to non-ignorable empirical missingness at some point in time. It would, therefore, be wise to include MNAR missingness in the simulation study just in case. Alternatively, a sensitivity analysis may provide an indication of the validity of the obtained inference, given that the assumed missingness mechanism is suspected to be invalid \citep[see e.g.][part 5]{molenberghs2014handbook}.

In addition to varying the missingness mechanisms, each simulated mechanism should be combined with different missingness patterns. Remember that missingness is only ignorable under MAR when the parameter of the data is distinct and a-priori independent from the parameter of the missing data process. Under MAR missingness we assume that we may use the observed data to make inferences about the joint (observed and unobserved) data. The dependency of the procedure on the assumption under which we obtain inference is only influenced by the amount of missingness. If there is no missingness--or if there is no data, for that matter--the inference does not depend on the assumption. Alternatively, the validity of assumptions becomes increasingly important when the missingness increases. Since we control the MAR mechanism, the assumption under which we may solve the missing data problem should hold and it is only fair to assess performance under stringent missingness conditions. We, therefore, propose to evaluate imputation methodology under 10\%, 25\% and 50\% univariate missingness (see Table \ref{table:prop}).
